---
title: "Transplantation study: supplementary analysis and plots"
author: "Giulia"
date: "2/2/2022"
output: html_document
---

#Load packages
```{r}
#R version in use: version.string R version 3.6.1 (2019-07-05)

#load pachages
library(ampvis2) #version 2.5.1
library(tidyverse) #version 1.2.1
library(ggplot2) #version now 3.3.5
```

#Fig.1 - pictures of transplantation

#Fig.S2-Overview of samples collected for microbial community analysis.

```{r}
#import long format
d_long_filt <- read.table(file = 'data_generated/d_long_filt.tsv', sep = '\t', header = TRUE, encoding = "UTF-8", stringsAsFactors = F)
```

```{r}
# Prepare metadata

## subset needed column
d_meta <- d_long_filt %>% 
          select(Plants, SampleType, Date, Days, Transplantation, SampleGroup) %>% 
          unique()

## add a simple grouping for the plot
d_meta$SampleGroup2[d_meta$SampleType %in% "IWW"] <- "IWW"
d_meta$SampleGroup2[d_meta$SampleGroup %in% "D-AS(before)" | d_meta$SampleGroup %in% "R-AS(before)"] <- "AS(before)"
d_meta$SampleGroup2[d_meta$SampleGroup %in% "D-AS(after-early)" | d_meta$SampleGroup %in% "R-AS(after-early)"] <- "AS(after-early)"
d_meta$SampleGroup2[d_meta$SampleGroup %in% "D-AS(after-intermediate)" | d_meta$SampleGroup %in% "R-AS(after-intermediate)"] <- "AS(after-intermediate)"
d_meta$SampleGroup2[d_meta$SampleGroup %in% "D-AS(after-late)" | d_meta$SampleGroup %in% "R-AS(after-late)"] <- "AS(after-late)"

## sort grouping
d_meta$SampleGroup2 <- factor(d_meta$SampleGroup2, levels = c("IWW",
                                                          "AS(before)",
                                                          "AS(after-early)",
                                                          "AS(after-intermediate)",
                                                          "AS(after-late)"))

# Plot
library(ggrepel)

p_meta <- ggplot(d_meta, aes(x = Days, y=SampleType, color = SampleGroup2)) +
                geom_point(alpha = 1) +
                geom_vline(xintercept = 0, linetype = "dashed") +
                geom_text(data = subset(d_meta, !Days %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8)), aes(label = Days), hjust = -0.5,  vjust = 0.2, angle = 90, size = 3, show.legend = FALSE) +
               geom_text_repel(data= subset(d_meta, Days %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8)), 
                            aes(label = Days,
                                segment.square = F,
                                segment.inflect = F),
                               show.legend = FALSE,
                                angle = 90, size = 3,
                                vjust = -0.5, 
                                direction         = "x",
                                segment.size      = 0.5,
                                min.segment.length = 0
                              )  +
 
                facet_wrap(~Plants, ncol = 1, nrow = 2) +
                ylab("Sample type") +
                theme_bw() +
                theme(legend.position = "right",
                      strip.text = element_text(face = "bold", size = 13),
                      axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
                scale_x_continuous(breaks=seq(-140,200,10)) +
                scale_color_manual(values = c("IWW" = "darkgrey",
                                                "AS(before)" = "black",
                                                "AS(after-early)" = "#8351DA",
                                              "AS(after-intermediate)" = "#2D949A",
                                              "AS(after-late)" = "#D05801"),
                                   name = "Sample Group") 
                

ggsave(plot = p_meta, filename = "C:/Users/Giulia/Documents/R/Immigration/Transplantation/paper_codes/Figures/Fig.S2-Samples overview.jpeg", width = 10, height = 4)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```

#Fig.S3 - DSVI donor AS
```{r}
d_dsvi <- readxl::read_excel("data_raw/Process parameters/SVI- 2014_2017_merged.xlsx")

```

```{r}
#subset
d_dsvi2 <- d_dsvi %>% 
  select(7, 8) %>% 
  magrittr::set_colnames(c("Days", "DSVI30")) %>% 
  filter(Days >= -500)

#plot
ggplot(d_dsvi2, aes(x = Days, y = DSVI30)) +
  geom_point(colour = "black", alpha = 0.8, size = 3) +
  geom_line(colour = "black", alpha = 0.4) +
  labs(x = "Days", y = "DSVI [mL/g]") +
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
  theme_bw() +
  scale_x_continuous(breaks=seq(-500,150,50)) +
scale_y_continuous(breaks=seq(0,300,50), limits = c(0, 300)) 

ggsave("Figures/Fig.S3 - DSVI D-AS.png", height = 4, width = 6)
```

#Fig.S4-Distance plots to D-AS(before) model.
(from statistial analysis, file "2-full analysis - 0.01 pct -v10-SVK-gd.Rmd")

#Fig.S5-Distance plots to R-AS(before) model.
(from statistial analysis, file "2-full analysis - 0.01 pct -v10-SVK-gd.Rmd")

#Fig.S6-Abundance variation over time before transplantation of selected genera.

###Import filtered dataset
```{r}
#import ampvis format
d_amp_filt <- readRDS('data_generated/d_amp_filt.rds')
```

###Subset 
```{r}
#Subset AS samples only
subset <- amp_subset_samples(d_amp_filt, SampleType %in% "Activated Sludge" & Transplantation %in% "Before", normalise = F) 

subset$metadata$Date <- as.factor(as.Date(subset$metadata$Date))
```

###Prepare dataset
```{r}
#create list of top 5 genus with abundances
p<-  amp_heatmap(subset,
            group_by = c("Days","Plants"),
            tax_aggregate = "Genus",
            tax_show = 5,
            normalise = F,
            color_vector = c("white", "red"),
            plot_colorscale = "log10",
            plot_values = TRUE,
            plot_values_size = 8) 


p_ab <- p$data

#check the top5 most abunda genera. They are the same in R-AS and D-AS
p_ab %>% 
  group_by(Plants, Display) %>% 
  summarise(mean_ab = mean(Abundance)) %>% 
   arrange(Plants, desc(mean_ab))

#export genera vector as ordered by abundance in donor
toptaxa <- p_ab %>% 
  group_by(Plants, Display) %>% 
  summarise(mean_ab = mean(Abundance)) %>% 
  arrange(Plants, desc(mean_ab)) %>% 
  subset(Plants %in% "Donor") %>% 
  pull(Display) %>% 
  as.character()


#create data frame with days, plant and top 15 genus with abundances
pp<- data.table::dcast(p$data, .Group~Display,value.var = "Abundance",fun.aggregate = mean)

#transpose data frame
ppmb1 <- gather(pp, key=Taxa, value = Abundance,-.Group)

#split days and plants column
ppmb2 <- ppmb1 %>% 
  separate(.Group, c("Days", "Plant"), sep = " ")

ppmb2$Days = as.numeric(as.character(ppmb2$Days))

#order taxa
ppmb2$Taxa <- factor(ppmb2$Taxa, levels = toptaxa)
```

###Plot
```{r}
#plot figure top row
p_gen <- ggplot(ppmb2, aes(x = Days, y = Abundance, color=Plant))+ 
              geom_point(size=3, alpha = 1) +
              geom_line(size=1, alpha = 0.4) +
                 theme(legend.position = "bottom",
            panel.background = element_blank(),
            axis.line.x = element_line(color = "black"),
            axis.line.y = element_line(color = "black"),
            legend.text = element_text(size = 24, color = "black"),
            legend.title = element_blank(),
            axis.text.y = element_text(size = 15, color = "black"),
            axis.text.x = element_text(size = 15, hjust = 1, vjust = 0.5, color = "black", angle = 90),
            axis.title.x = element_text(size = 20, color = "black"),
            axis.title.y = element_text(size = 20, color = "black"),
            strip.text = element_text(size = 24, color = "black", face = "bold"),
            plot.title = element_text(size = 28, color = "black"),
            plot.subtitle = element_text(size = 26, color = "black"),
            ) +
         
  facet_wrap(~ as.factor(Taxa), scales = "free_y", ncol = 5)+
  xlab("Days") + 
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1)+
  ylab("Relative read abundance [%]") +
  scale_x_continuous(breaks=seq(-140,220,20)) +
  scale_color_manual(labels = c("D-AS", "R-AS"),
                       values = c("#CEAB42", "#B560BF"))  

# ggsave(plot = p_gen, "~/R/Immigration/Transplantation/paper_codes/Figures/Fig.S6 - Top 5 genera before transplantation.png", height = 5, width = 20)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```

#Fig.S7 - Heatmap of IWW before transplantation in donor and recipient

```{r}
#import ampvis format
d_amp_filt <- readRDS('data_generated/d_amp_filt.rds')
```

```{r}
iww_before <- amp_subset_samples(d_amp_filt, SampleType %in% "Wastewater influent" & Transplantation %in% "Before")

## Add a new column to use for the faceting of the heatmap 
iww_before$metadata$PlantSampleType[iww_before$metadata$Plants %in% "Recipient" & iww_before$metadata$SampleType %in% "Wastewater influent"] <- "R-IWW"
iww_before$metadata$PlantSampleType[iww_before$metadata$Plants %in% "Donor" & iww_before$metadata$SampleType %in% "Wastewater influent"] <- "D-IWW"
```

```{r}
amp_heatmap(iww_before,
            group_by = "Days",
            facet_by = c("Plants", "Transplantation"),
            normalise = F,
            tax_aggregate = "Species",
            tax_add = "Genus",
            tax_show = 50)
```

HM sorted by donor
```{r}
#creating hm to get the list
iww_don_bef <- amp_subset_samples(iww_before, SampleType %in% "Wastewater influent" & Transplantation %in% "Before" &Plants %in% "Donor")

top50spDonor <- amp_heatmap(iww_don_bef,
                            group_by = "Days",
            #facet_by = c("Plants", "Transplantation"),
            facet_by = c("PlantSampleType"),
            normalise = F,
            tax_aggregate = "Species",
            tax_add = "Genus",
            tax_show = 50,
            plot_colorscale = "log10", 
            #sort_by = "16SAMP-4790",
            round = 2,
           textmap = T)

#extract the list of species
top50spDonor <- rownames(top50spDonor)

#HM showing list
iww_before$metadata$SampleType[iww_before$metadata$SampleType == "Wastewater influent"] <- "IWW"

plottest <- amp_heatmap(iww_before,
                         group_by = "Transplantation",
           # facet_by = c("Plants", "SampleType", "Transplantation"),
            facet_by = c("PlantSampleType"),
            normalise = F,
            tax_aggregate = "Species",
            tax_add = "Genus",
            tax_show = top50spDonor,
            plot_colorscale = "log10", 
            #sort_by = "16SAMP-4790",
            round = 2)


plottest

#reorder data

head(plottest$data)
unique(plottest$data[,2])

plottest$data[,2] <- as.character(plottest$data[,2]) #instead of droplevels() as there may be different taxa between 
plottest$data[,2] <- factor(plottest$data[,2], levels = rev(top50spDonor))


#sorted plot

plottest +
  labs(title = "Before") +
  theme(axis.text.y = element_text(size = 12, colour = "black"), 
        axis.text.x = element_text(size = 12, colour = "white", vjust = 0.5, hjust = 0.5, angle = 0), 
        legend.position = "right",
        strip.text = element_text(size = 13, face = "bold", color = "black"),
         plot.title = element_text(size = 15, color = "black", hjust = 0.5)
             ) 

#ggsave(filename = "Figures/Fig.S7-HM_IWW before_sorted by donor_top50species.pdf", width = 10, height = 8, dpi = 600)
ggsave(filename = "~/R/Immigration/Transplantation/paper_codes/Figures/Fig.S7-HM_IWW before_sorted by donor_top50species.pdf", width = 10, height = 8, dpi = 600)
```

#Fig.S8-Sludge retention time of recipient plant.

###Import raw metadata

```{r Metadata Sludge}
### Sludge ###
Data <- read.delim(file = "data_raw/Process parameters/NØ_Slamrapport_Combined.txt", sep = "\t", header = T, dec = ".", encoding = "UTF-16", na.strings = "NULL") 

names(Data)[names(Data) == "Date"] <- "StartTime"

Sludge <-  reshape2::melt(Data, id = c("StartTime"), measure.vars = colnames(Data)[2:ncol(Data)], variable.name = "Parameter", value.name = "Value") #3 gives the starting column number for column names to store as measure variables. This differs depending on the original dataformat.
Sludge$Origin <- "Process"

Sludge$Value <- as.numeric(Sludge$Value)

Sludge$Days <- lubridate::as_date(Sludge$StartTime, "%Y-%B-%d") - lubridate::as_date("2017-04-20", "%Y-%B-%d") #calculate Days of transplantation considering 2017-0420 = Day0

Sludge$StartTime <- strptime(Sludge$StartTime, format = c("%Y-%m-%d"))
#The POSIXlt format differs between original dataformats...

Sludge_Days <- Sludge
Sludge_Days$StartTime <- trunc(Sludge$StartTime, "days") %>%
as.character() %>%
as.factor()

Sludge_Days <- aggregate(Value~StartTime+Parameter+Origin+Days, data = Sludge_Days, FUN = "mean")

Sludge_Days$StartTime <- strptime(Sludge_Days$StartTime, format = c("%Y-%m-%d"))
```

###Sludge age calculation
```{r}
#Sludge Age
WAS_MASS <- subset(Sludge, Parameter %in% "WAS_Mass")
Total_Biomass <- subset(Sludge, Parameter %in% "LT2_TS")
Total_Biomass <- select(Total_Biomass, -Origin)
Total_Biomass$Parameter <- "Total Sludge Age"
SludgeAge <- Total_Biomass
SludgeAge$Value <- (Total_Biomass$Value*2*1750)/WAS_MASS$Value
SludgeAge <- subset(SludgeAge, Value < 50)
```

###Plot

```{r}
###Timeseries
p_timeseries <- ggplot(SludgeAge, aes(x = Days, y = as.numeric(Value))) +
   geom_point(size = 1.8, alpha = 0.5) +
    geom_line(size = 1.0, alpha = 0.2) +
    theme(legend.position = "none",
           panel.background = element_blank(),
           legend.key = element_rect(fill="transparent"),
           legend.text = element_text(size = 12, face = "bold", color = "black"),
           legend.title = element_blank(),
           axis.text.y = element_text(size = 11, color = "black"), 
           axis.text.x = element_text(size = 11, color = "black", hjust = 0.5, vjust = 0.5, angle = 90), 
           axis.title.x  = element_text(size = 12, color = "black"),
           axis.title.y  = element_text(size = 12, color = "black"),
           
           panel.grid.major = element_line(size = 0.2, linetype = 'solid', colour = "gray"),
           panel.grid.minor = element_blank(),
           strip.background = element_blank(),
           panel.border = element_rect(colour = "black", fill = NA),
           strip.text = element_text(size = 15, face = "bold", color = "black", hjust = 0)
           ) +
   xlab("Days") +
  ylab("Total Sludge Retention Time [d]") +
   geom_smooth(size = 2, se = FALSE, span = 0.1, color = "red") +
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 2) +
  geom_vline(xintercept = -140, linetype = "longdash", color = "black", size = 1) +
   geom_vline(xintercept = 203, linetype = "longdash", color = "black", size = 1) +
   scale_x_continuous(breaks=seq(-500,200,50)) +
   scale_y_continuous(breaks = c(0,5,10,15,20,25,30)) 


###boxplot

#add a column to identify the days before and after transplantation that correpond to the days which the community data is available for
SludgeAge$Transplantation[SludgeAge$Days >= -140 & SludgeAge$Days <= 0] <- "Before"
SludgeAge$Transplantation[SludgeAge$Days > 0 & SludgeAge$Days <= 203] <- "After"

SludgeAge_subset <- subset(SludgeAge, !is.na(Transplantation))

SludgeAge_subset$Transplantation <- factor(SludgeAge_subset$Transplantation, levels = c("Before", "After"))

p_boxplot <- ggplot(SludgeAge_subset, aes(x=Transplantation, y=Value)) + 
  geom_boxplot() +
   theme(legend.position = "none",
           panel.background = element_blank(),
           legend.key = element_rect(fill="transparent"),
           legend.text = element_text(size = 12, face = "bold", color = "black"),
           legend.title = element_blank(),
           axis.text.y = element_text(size = 11, color = "black"), 
           axis.text.x = element_text(size = 11, color = "black", hjust = 0.5, vjust = 0.5, angle = 0), 
           axis.title.x  = element_text(size = 12, color = "black"),
           axis.title.y  = element_text(size = 12, color = "black"),
          panel.grid.major = element_line(size = 0.2, linetype = 'solid', colour = "gray"),
         strip.background = element_blank(),
           panel.border = element_rect(colour = "black", fill = NA),
           ) +
   ylab("Total Sludge Retention Time [d]") +
  scale_y_continuous(breaks = seq(0,30,5)) 

###combine plots

library(ggpubr)

p_srt <- ggarrange(p_timeseries, NULL, p_boxplot,
                     labels = c("A)", "", "B)"),
                    font.label = list(size = 15, color = "black", face = "bold", family = NULL),
                    ncol = 3,
                    nrow = 1,
                    widths = c(1, 0.05, 0.3)
                ) 

ggsave(plot = p_srt, "Figures/Fig.S4-SRT recipient.jpg", width = 10, height = 4)
```
Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```

#Fig.S9-Possible responses of microbial community to transplantation.
Theory image made in Inkscape

#Fig.S10-Removal of abundant species from recipient through transplantation.  

###Import filtered dataset
```{r}
#import long format
d_long_filt <- read.table(file = 'data_generated/d_long_filt.tsv', sep = '\t', header = TRUE, encoding = "UTF-8", stringsAsFactors = F)
```

###Import Venn-dataframe
```{r}
#import dataframe from Venn
df_venn <- readxl::read_excel("data_generated/df_venn.xlsx")
```

###Subset
```{r}
#select unique in recipient
ASR_before_unique <- subset(df_venn, VennGroup %in% "ASR_before_unique")

#subset unique species of recipient from main dataset
rec <- subset(d_long_filt, Plants %in% "Recipient" & SampleType %in% "AS" &
                Species_rename %in% ASR_before_unique$Species_rename)

#subset species before transplantation with mean abundance >= 0.1%
rec3_ab <- rec %>% 
  filter(Transplantation %in% "Before") %>% 
  group_by(Species_rename) %>% 
   summarise(rra_avg = mean(rra_sp)) %>% 
  filter(rra_avg >= 0.1)

unique(rec3_ab$Species_rename)

#subset them from main dataset and in R-AS
rec_sub <- subset(rec, Species_rename %in% rec3_ab$Species_rename)
```

##Top 5 species
```{r}
#sort by abundance in R-AS (1st day after transplantation) and extract top 5
top5 <- subset(rec_sub, Days %in% "1") %>% 
            select(Species_rename2, rra_sp) %>% 
            arrange(desc(rra_sp)) %>% 
            head(., 5) %>% 
            pull(Species_rename2) %>% 
            as.character()
     
#subset from full dataset with donor
both_top5 <- subset(d_long_filt, SampleType %in% "AS" & Species_rename2 %in% top5)

both_top5$Species_rename2 <- factor(both_top5$Species_rename2, levels = top5)
```

##Plot
```{r}
p_removed <- ggplot(both_top5, aes(x = Days, y = rra_sp, color=Plants, group = Plants)) + 
                geom_point(size=3, alpha = 0.8) +
              geom_line(size=1, alpha = 0.4) +
              theme(legend.position = "bottom",
                         panel.background = element_blank(),
                         axis.line.x = element_line(color = "black"),
                         axis.line.y = element_line(color = "black"),
                         legend.text = element_text(size = 15, color = "black"),
                         legend.title = element_blank(),
                         axis.text.y = element_text(size = 13, color = "black"), 
                         axis.text.x = element_text(size = 13, angle = 90, vjust = 0.5, hjust = 1, color = "black"), 
                         axis.title.x = element_text(size = 15, color = "black"),
                         axis.title.y = element_text(size = 15,color = "black"),
                         strip.text = element_text(size = 18, face = "bold", color = "black"),
                         plot.title = element_text(size=25),
                         plot.subtitle = element_text(size = 20)
                         ) +
              facet_wrap(~ Species_rename2, scales = "free_y", ncol = 5, labeller = label_wrap_gen(width=3))+
                 xlab("Days") + 
                geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
                ylab("Relative read abundance [%]") + 
                scale_x_continuous(breaks=seq(-140,220,40)) +
                 scale_color_manual(labels = c("D-AS", "R-AS"),
                                    values = c("#CEAB42", "#B560BF"))  

# ggsave(plot = p_removed, "~/R/Immigration/Transplantation/paper_codes/Figures/Fig.S10-Removed top5.png", height = 4, width = 18)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```

#Fig.S11-Effect of transplantation for PAOs at species level.

## Import filtered dataset
```{r}
#import long format
d_long_filt <- read.table(file = 'data_generated/d_long_filt.tsv', sep = '\t', header = TRUE, encoding = "UTF-8", stringsAsFactors = F)
```

## Import venn-diagram species
```{r}
#import dataframe from Venn to get shared species before
df_venn <- readxl::read_excel("data_generated/df_venn.xlsx")

#subset for species shared in D-AS and R-AS before transplantation (528 species)
sb <- subset(df_venn, VennGroup %in% "NA_before_shared")

## Subset shared species from long format dataset
shared <- subset(d_long_filt, SampleType %in% "AS" & Species_rename %in% sb$Species_rename)
#unique(shared$Species_rename) #check that total number of subsetted species is 528
```

## Import guilds
```{r}
guilds <- readxl::read_excel("data_raw/functional guilds MiDAS3.6.xlsx")

# Data wrangling of guilds names
guilds <- guilds %>% 
  drop_na() %>% 
   subset(., !Guild %in% "potential filament in WWTP") #not important

guilds_gendf <- guilds %>% 
  filter(!`New name`%in% c("midas_s_328", "Ca_Defluviicoccus_seviourii")) %>% #exclude guilds species
  rename(Genus = `New name`)
guilds_gendf$Genus <- paste0("g__", guilds_gendf$Genus) #46 genera
guilds_gen <- guilds_gendf$Genus

guilds_spdf <- guilds %>% 
  filter(`New name`%in% c("midas_s_328", "Ca_Defluviicoccus_seviourii")) %>% #include guilds species
  rename(Species_rename = `New name`)
guilds_spdf$Species_rename <- paste0("s__", guilds_spdf$Species_rename) #2 species
guilds_sp <- guilds_spdf$Species_rename
```

```{r}
# Subset species belonging to guilds from long format dataset
guilds_sub <- subset(d_long_filt, SampleType %in% "AS" & (Genus %in% guilds_gen | Species_rename %in% guilds_sp))
unique(guilds_sub$Genus)

# Subset shared species in D-AS and R-AS before transplantation from guilds
guilds_shared <- subset(guilds_sub, Species_rename %in% unique(shared$Species_rename))
unique(guilds_shared$Species_rename) #(53 species)

## attach guild tag to the shared guilds before transplantation, by merging with guild genus dataset
guilds_shared2 <- merge(guilds_shared, guilds_gendf, by = "Genus") #merge with genus
unique(guilds_shared2$Species_rename)
```


##Subset
```{r}
#subset for PAOs
pao <- guilds_shared2 %>% filter(Guild %in% "PAO") 

#Stats: how many species in each genera belonging to guilds
pao %>% 
  group_by(Guild, Genus) %>% 
  distinct(Species_rename, .keep_all = T) %>% 
  dplyr::summarise(nrofspecies = n())

#exclude species with average rra less than 0.01 in both donor and recipient AS
pao %>% 
  filter(Plants %in% "Recipient" & Transplantation %in% "Before") %>% #for R-AS only "before" transplantation is representative of steady state
  group_by(Plants, Transplantation, Species_rename) %>%
  dplyr::summarise(mean = mean(rra_sp)) %>% 
  arrange(desc(mean)) %>% 
  filter(mean < 0.01)
#in R-AS all PAO species are > 0.01 on average before transplantation   
   
pao %>% 
    filter(Plants %in% "Donor") %>%   #for D-AS both "before" and "after" transplantation is representative of steady state, so the whole sampling period is consider to check the average rra
   group_by(Plants, Species_rename) %>%
    summarise(mean = mean(rra_sp)) %>% 
  arrange(desc(mean)) %>% 
  filter(mean < 0.01) #%>% 
 # pull(Species_rename)

pao2 <- pao %>% 
    filter(!Species_rename %in% c("s__midas_s_1177", "s__midas_s_1827"))

#exclude PAO genera with less than 2 species - exlcude Gemmatimonas genus. Not useful to see how different species within a genus behave.
pao3 <- pao2 %>% filter(!Genus %in% "g__Gemmatimonas")

#create a column with only species or ASV name
pao4 <- pao3 %>% 
  mutate(Species_rename3 = paste(Species_rename2)) %>% 
  separate(Species_rename3, c("Gen", "SpeciesASV"), sep = " ") %>% 
  select(-Gen)

#Create a column with tidy Genus name
pao4 <- pao4 %>% 
  mutate(Genus2B = paste(Genus)) 
pao4$Genus2B <- gsub("g__", "", pao4$Genus2B)
```

##Plot (loop)
```{r}
#plot species within PAOs genera

gg_PAO <- list()#list to store plots

#plot
for (i in unique(pao4$Genus2B)) {
  
  sub <- subset(pao4, Genus2B == i)

gg_PAO[[i]] <- ggplot(sub, aes(x = Days, y = rra_sp, color=Plants)) + 
  geom_point(size=3, alpha = 0.8) +
  geom_line(size=1, alpha = 0.4)+
      theme(legend.position = "bottom",
           panel.background = element_blank(),
           axis.line.x = element_line(color = "black"),
           axis.line.y = element_line(color = "black"),
           legend.text = element_text(size = 18, face = "bold", color = "black"),
           legend.title = element_blank(),
           axis.text.y = element_text(size = 12, color = "black"), 
           axis.text.x = element_text(size = 12, color = "black", angle = 90, vjust = 0.5, hjust = 1), 
           axis.title.x = element_text(size = 15, color = "black"),
           axis.title.y = element_text(size = 15, color = "black"),
           strip.text = element_text(size = 18, face = "bold", color = "black"),
           plot.title = element_text(size=25, hjust = 0.5),
           plot.subtitle = element_text(size = 20)
           ) +
  facet_wrap(~ SpeciesASV, scales = "free", ncol = 5, labeller = label_wrap_gen(width=3))+
  labs(title = paste0("",i,"")) +
  xlab("Days") + 
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
  ylab("Relative read abundance [%]") + 
  scale_color_manual(labels = c("D-AS", "R-AS"),
                        values = c("#CEAB42", "#B560BF"))  +
  scale_x_continuous(breaks=seq(-140,220,40)) 
 
 }


# sort species within each genus
gg_PAO$Tetrasphaera$data$SpeciesASV <- factor(gg_PAO$Tetrasphaera$data$SpeciesASV, levels = c("midas_s_5", "midas_s_45", "midas_s_328", "Tetrasphaera_elongata", "midas_s_1378"))

gg_PAO$Ca_Accumulibacter$data$SpeciesASV <- factor(gg_PAO$Ca_Accumulibacter$data$SpeciesASV, levels = c("ASV154", "Ca_Accumulibacter_phosphatis", "midas_s_315", "midas_s_2081"))

gg_PAO$Tessaracoccus$data$SpeciesASV <- factor(gg_PAO$Tessaracoccus$data$SpeciesASV, levels = c("midas_s_307", "midas_s_413", "midas_s_1151"))

gg_PAO$Dechloromonas$data$SpeciesASV <- factor(gg_PAO$Dechloromonas$data$SpeciesASV, levels = c("midas_s_173", "ASV85"))

```

##Combined plot 
```{r}
library(cowplot)

tessdech <- plot_grid(gg_PAO$Tessaracoccus + theme(legend.position="none"), 
                      gg_PAO$Dechloromonas + theme(legend.position="none"),
                      labels = c("C)", "D)"),
                      label_size = 20,
                      ncol = 2,
                      nrow = 1,
                      rel_widths = c(1.5, 1))

pall <- plot_grid(gg_PAO$Tetrasphaera + theme(legend.position="none"),
                  gg_PAO$Ca_Accumulibacter + theme(legend.position="none"),
                  tessdech,
                  labels = c("A)", "B)", ""),
                  label_size = 20,
                  ncol = 1,
                  nrow = 3,
                   align = 'vh',
                  rel_widths = c(1.5, 1)
                  ) 

#get legend
# extract a legend that is laid out horizontally
legend <- get_legend(
            gg_PAO$Tetrasphaera + 
            theme(legend.position = "bottom")
            )

# add the legend underneath the row we made earlier. Give it 5% of the height of one plot (via rel_heights).
pall2 <- plot_grid(pall, 
                  legend, 
                  ncol = 1, 
                  rel_heights = c(1, .05)
                  ) 

#ggsave(plot = pall2, "~/R/Immigration/Transplantation/paper_codes/Figures/Fig.S11-PAOs.png", width = 18, height = 12)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```

#Fig.S12-Effect of transplantation for genera belonging to functional guilds.
###Import filtered dataset
```{r}
#import long format
d_long_filt <- read.table(file = 'data_generated/d_long_filt.tsv', sep = '\t', header = TRUE, encoding = "UTF-8", stringsAsFactors = F)
```

###Genera of interest
```{r}
pao_target <- c("g__Tetrasphaera","g__Ca_Accumulibacter", "g__Dechloromonas", "g__Tessaracoccus")
gao_target <- c("g__Micropruina","g__Defluviicoccus", "g__Ca_Competibacter", "g__Ca_Contendobacter")
nitrifiers_target <- c("g__Nitrosomonas", "g__Nitrotoga", "g__Nitrospira")
filaments_target <- c("g__Ca_Microthrix","g__Trichococcus", "g__Ca_Promineofilum", "g__Ca_Villigracilis", "g__Ca_Amarolinea", 
               "g__Gordonia", "g__Thiothrix", "g__Ca_Sarcinithrix", "g__Leptothrix", "g__Sphaerotilus")

```

###Subset
subset genera from main dataset 
```{r}
as <- subset(d_long_filt, SampleType %in% "AS")
```

###Abundance at genus level
```{r}
#recalculate abundance at genus level
as_gen <- as %>%
  group_by(Plants, SampleType, Days, Genus) %>% 
  mutate(rra_gen = sum(rra_sp)) %>% 
  dplyr::select(-Genus2, -Species_rename, -Species_rename2, -Taxa, -SampleReads, -count_sp, -rra_sp) %>% #remove all columns that contain info at species level 
  unique()

#remove g__ from Genus name
as_gen$Genus2 <- gsub("g__", "", as_gen$Genus)
```

###Subset guilds
```{r}
#subset from main dataset
pao <- subset(as_gen, Genus %in% pao_target) %>% mutate(., Guild = "PAOs")
gao <- subset(as_gen, Genus %in% gao_target) %>% mutate(., Guild = "GAOs")
nit <- subset(as_gen, Genus %in% nitrifiers_target) %>% mutate(., Guild = "Nitrifiers")
fil <- subset(as_gen, Genus %in% filaments_target) %>% mutate(., Guild = "Filamentous bacteria")

#merge guilds datasets into a dataframe
guilds_df <- bind_rows(pao, gao, nit, fil)
```

###Plot guilds (loop)
```{r}
plot <- c()

for(i in unique(guilds_df$Guild)) {
  
  #subset the guild
  sub <- subset(guilds_df, Guild == i)
  
  #extract the genera for faceting
  target <- sub %>% 
  filter(Plants %in% "Donor") %>% 
  group_by(Genus2) %>% 
  summarise(mean = mean(rra_gen)) %>% 
  arrange(desc(mean)) %>% 
  pull(Genus2) 
  
  #set genera as levels
  sub$Genus2 <- factor(sub$Genus2, levels = target)
  
  #plot guild
  plot[[i]] <- ggplot(sub, aes(x = Days, y = rra_gen, color=Plants)) + 
  geom_point(size=3, alpha = 1) +
  geom_line(size=1, alpha = 0.4)+
     theme(legend.position = "bottom",
           panel.background = element_blank(),
           axis.line.x = element_line(color = "black"),
           axis.line.y = element_line(color = "black"),
           legend.text = element_text(size = 18, face = "bold", color = "black"),
           legend.title = element_blank(),
           axis.text.y = element_text(size = 12, color = "black"), 
           axis.text.x = element_text(size = 12, color = "black", angle = 90, vjust = 0.5, hjust = 1), 
           axis.title.x = element_text(size = 15, color = "black"),
           axis.title.y = element_text(size = 15, color = "black"),
           strip.text = element_text(size = 18, face = "bold", color = "black"),
            plot.title = element_text(size=23, hjust = 0.5),
           ) +
  facet_wrap(~ Genus2, scales = "free", ncol = ifelse(i == "PAOs", 4,
                                                 ifelse(i == "GAOs", 4,
                                                    ifelse(i == "Nitrifiers", 3,
                                                        ifelse(i == "Filamentous bacteria", 5,
                                                        NA))))
                          )+
  xlab("Days") + 
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
  ylab("Relative read abundance [%]") + 
  scale_x_continuous(breaks=seq(-140,220,40)) +
  scale_color_manual(labels = c("D-AS", "R-AS"),
                        values = c("#CEAB42", "#B560BF"))  +
  labs(title = paste0(i)) 
}

#extract legend
p_legend <- plot$PAOs
```

###Combined plot
```{r}
library(ggpubr)

legend <- get_legend(p_legend)
#as_ggplot(legend)

p_guilds_all <- ggarrange(plot$PAOs + rremove("legend"), 
                          plot$GAOs + rremove("legend"), 
                          plot$Nitrifiers + rremove("legend"), 
                          plot$`Filamentous bacteria` + rremove("legend"), 
                          legend,
          ncol = 1,
          nrow = 5,
           labels = c("A)", "B)", "C)", "D)"),
           font.label = list(size = 23, face = "bold"),
             heights = c(1, 1, 1, 1.5, 0.1)) 

# ggsave(plot = p_guilds_all, "~/R/Immigration/Transplantation/paper_codes/Figures/Fig.S12-Time-series genera guilds.png", height = 16, width = 20)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```


#Fig.13-Variation over time of abundance of selected species in AS and IWW from the recipient plant. 

###Import filtered dataset
```{r}
#import long format
d_long_filt <- read.table(file = 'data_generated/d_long_filt.tsv', sep = '\t', header = TRUE, encoding = "UTF-8", stringsAsFactors = F)
```

##Select species (as.Fig.8)
```{r}
#A)Species from model distances. 
##Select species that contribute the most to the similarity between R-IWW/R-AS(before) and R-IWW/R-AS(after-late) and have a high VIP score. 
A <- c("f__Ruminococcaceae__ASV3228",
        "s__midas_s_299",
       "s__midas_s_2082",
       "f__Sphingomonadaceae__ASV35937",
       "s__midas_s_207"
                )

#B)Added species.
##Selected unique and abundant species of D-AS added to R-AS
B <- c("s__midas_s_1",
       "g__Dietzia__ASV97",
       "s__midas_s_243",
       "s__Sphingopyxis_bauzanensis",
       "s__midas_s_220")

#C)Species within guilds
C <- c("s__Ca_Microthrix_parvicella",
       "s__midas_s_403",
       "s__midas_s_4",
       "s__midas_s_307",
       "s__midas_s_5")
```

###Subset
```{r}
#subset for recipient AS samples 
recz <- subset(d_long_filt, Plants %in% "Recipient") 

# Subset the 3 groups of species from recipient dataset
rec_A <- subset(recz, Species_rename %in% A) %>% mutate(., SpGroup = "A")
rec_B <- subset(recz, Species_rename %in% B) %>% mutate(., SpGroup = "B")
rec_C <- subset(recz, Species_rename %in% C) %>% mutate(., SpGroup = "C")

#merge species group into a dataframe
rec_abc <- bind_rows(rec_A, rec_B, rec_C)
```

###Plots (loop)
```{r}
#plot
plot <- list()

for(i in unique(rec_abc$SpGroup)) {
  
  #subste for each species group (A, B, C)
  sub <- subset(rec_abc, SpGroup == i)
  
  for(y in unique(sub$SampleType)) {
    
    #subste for each species type (AS or IWW)
    sub2 <- subset(sub, SampleType == y)
    
    #plot
    plot[[i]][[y]] <- ggplot(sub2, aes(x = Days, y = rra_sp, group = Species_rename2)) + 
                  geom_point(size=3, color = ifelse(sub2$SampleType == "AS", 
                                                    "darkblue",
                                                    "#4DB3E6")) +
                  geom_line(size=1, color = ifelse(sub2$SampleType == "AS", 
                                                    "darkblue",
                                                    "#4DB3E6"), alpha =0.6)+
                     theme(legend.position = "none",
                           panel.background = element_blank(),
                           axis.line.x = element_line(color = "black"),
                           axis.line.y = element_line(color = "black"),
                           legend.text = element_text(size = 10, color = "black"),
                           legend.title = element_blank(),
                           axis.text.y = element_text(size = 13, color = "black"), 
                           axis.text.x = element_text(size = 13, angle = 90, vjust = 0.5, color = "black"), 
                           axis.title.x = element_text(size = 15, color = "black"),
                           axis.title.y = element_text(size = 15,color = "black"),
                           strip.text = element_text(size = 18, face = "bold", color = "black"),
                           plot.title = element_text(size=25),
                           plot.subtitle = element_text(size = 20)
                           ) +
  facet_wrap(~ Species_rename2, scales = "free_y", ncol = 5, labeller = label_wrap_gen(width=3))+
   ylab("") +
   geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
  scale_x_continuous(limits = c(-140, 220), breaks=seq(-140,220,40))
    
  }
  
}

#impose factor for group A
levels_A <- c("Ruminococcaceae ASV3228", "Tetrasphaera midas_s_299", "Hyphomicrobium midas_s_2082", "Sphingomonadaceae ASV35937", "midas_g_207 midas_s_207")

plot$A$AS$data$Species_rename2 <- factor(plot$A$AS$data$Species_rename2, levels = levels_A)
plot$A$IWW$data$Species_rename2 <- factor(plot$A$IWW$data$Species_rename2, levels = levels_A)

#extract legend
p_legend <- ggplot(rec_A, aes(x = Days, y = rra_sp, color=SampleType, group =  interaction(SampleType, Species_rename2))) + 
  geom_point(size=3) +
  geom_line(size=1)+
     theme(legend.position = "bottom",
           panel.background = element_blank(),
           axis.line.x = element_line(color = "black"),
           axis.line.y = element_line(color = "black"),
           legend.text = element_text(size = 18, color = "black"),
           legend.title = element_blank(),
           axis.text.y = element_text(size = 13, color = "black"), 
           axis.text.x = element_text(size = 13, angle = 90, vjust = 0.5, color = "black"), 
           axis.title.x = element_text(size = 15, color = "black"),
           axis.title.y = element_text(size = 15,color = "black"),
           strip.text = element_text(size = 18, face = "bold", color = "black"),
           plot.title = element_text(size=25),
           plot.subtitle = element_text(size = 20)
           ) +
  facet_wrap(~ Species_rename2, scales = "free_y", ncol = 5, labeller = label_wrap_gen(width=3))+
  geom_vline(xintercept = 0, linetype = "longdash", color = "black", size = 1) +
  scale_x_continuous(breaks=seq(-140,220,40)) +
   scale_colour_manual(values = c("darkblue", "#4DB3E6"),
                       labels = c("R-AS", "R-IWW"))
```

###Extract each plot (loop)
```{r}
#combining AS and IWW from the sample species group into one plot
p_sub2 <- c()

for(i in unique(rec_abc$SpGroup)){
  
  p_sub <- plot[[i]]
  
  p_sub2[[i]] <- ggpubr::ggarrange(p_sub$AS + rremove("xlab"),
                                   p_sub$IWW + rremove("xlab"),
                                   ncol = 1,
                                   nrow = 2,
                                   align = "hv") 
  
  p_sub2[[i]] <- annotate_figure(p_sub2[[i]],
                left = text_grob("Relative read abundance [%]", size = 18, rot = 90),
                bottom = text_grob("Days", size = 18),
                fig.lab = ifelse(i == "A", "A)",
                                 ifelse(i == "B", "B)",
                                        ifelse(i == "C", "C)",
                                        "NA"))), 
                fig.lab.face = "bold",
                fig.lab.size = 25,
                fig.lab.pos = "top.left")
}

```

###Save combined plot
```{r}
library(ggpubr)

#get legend
legend <- get_legend(p_legend)
as_ggplot(legend)

#plot and legend together
p_all <- ggarrange(p_sub2$A, p_sub2$B, p_sub2$C, legend,
                ncol = 1,
                nrow = 4,
                heights = c(1, 1, 1, 0.1))

ggsave(plot = p_all, "Figures/Fig.S9-Time-series IWW and AS.jpeg", height = 16, width = 20)
```

Clean Environment for next analysis
```{r}
rm(list=ls(all=TRUE))
```


